# ->Analysis of Storm Damage<-
# ->(Reproducible Research:  Peer Assessment 2)<-


## Synopsis


## Data Processing

The data are read from the 'StormsData.csv' file. Column names are included n the file. Store data 'as-is', that is, do not do any implicit conversions, such as to factors.

```{r load_data,message=FALSE}
require(dplyr)
require(ggplot2)
require(lubridate)

storms <- read.table('StormData.csv',header=TRUE,sep=',',na.strings='',
                     ,as.is=TRUE,nrow=1000)
```

Print the beginning of the data.

```{r print1}
head(storms)
```

Many END_DATE entries are missing; if they are, then set them to the corresponding BGN_DATE. Then convert the begin date (BGN_DATE) and end date (END_DATE) fields to POSIXct class variables. It would be desirable to combine the DATE and TIME fields into a single field, but there are too many inconsistencies in the format of the time fields, and all this has not yet been investigated.

```{r convert_date}
# the following simulates ifelse(!is.na(END_DATE),END_DATE,BGN_DATE) which doesn't work ?
t <- storms$BGN_DATE
t[!is.na(storms$END_DATE)] <- storms$END_DATE[!is.na(storms$END_DATE)]
storms$END_DATE <- t

storms$BGN_DATE <- as.Date(storms$BGN_DATE,'%m/%d/%Y %H:%M:%S')
storms$END_DATE <- as.Date(storms$END_DATE,'%m/%d/%Y %H:%M:%S')

# would like to combine the start date and time, but there are too many
#  inconstencies
#update(storms$BGN_DATE,
#       hours=floor(as.numeric(storms$BGN_TIME)/100),
#       minute=as.numeric(storms$BGN_TIME)%%100)


```
## Results